Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-12-08 survey/survey.tex
Data I

\documentclass{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}

@@

@@

@@

\jc{wavefront}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data I: Finite Sets}

Most programming languages provide a rich collection of primitive
data like booleans, characters, strings, and numbers that are
naturally modeled as sets. Infinite sets embody a notion of recursion
which is subtle and is discussed briefly in the concluding
section. For finite sets, a reversible process between them is simply
a permutation. In other words, reversible programming with finite sets
is all about writing permutations.

We therefore start with a small language for writing permutations as a
product of transpositions:

\begin{verbatim}
p ::= id | swap i j | p;p
\end{verbatim}

Permutations on finite sets of size $n$ can be inductively defined as
follows:
\begin{verbatim}
perms(0)   = { id }
perms(n+1) = { (swap 0 i).p | p <- perms(n); i <- [1..n] }  
\end{verbatim}

>>>>

\documentclass{article}
\usepackage{amssymb,amsthm}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}

@@

\newtheorem{defn}{Definition}[section]

@@

\newcommand{\fin}[1]{\ensuremath{\left[#1\right]}}
\newcommand{\Nat}{\ensuremath{\mathbb{N}}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data I: Finite Sets}

Before we start our investigation of what it means to have typed,
information preserving transformations, let's begin with just
finite sets. Most programming languages provide some collection of primitive
data like booleans, characters, strings, and (bounded) numbers that are
naturally modeled as sets. Infinite sets are much more subtle, and will be
discussed in a later section.

What does it mean for two finite sets to be ``the same''?  Well, clearly the
sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$ are different.
Why?  Well, supposed there was a transformation $f : A \rightarrow B$ that
deformed $A$ into $B$, and another $g : B \rightarrow A$ which undid this
transformation. Since $f$ is total, by the pigeonhole principle, two elements
of $A$ would be mapped to the same element of $B$. Supposed that this is $2$ and
$3$, and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$, and
so $g$ is not the inverse of $f$. With just a little more work, we can show
that $f$ (and $g$) must be both injective and surjective. In other words,
$f$ (and $g$) must be a bijection between $A$ and $B$.  And of course this 
only happens when $A$ and $B$ have the same number of elements.

More importantly, given a bijection $f : C \rightarrow D$ of finite sets
$C,D$, there always exists another bijection $g : D \rightarrow C$ which is
$f$'s inverse. So, for finite sets, \emph{bijections} act as reversible
deformations.

This discussion is purely ``semantic'', in the sense that it is about
the denotation of simple primitive data (sets) and their reversible
deformations (bijections).  We would like to reverse engineer a programming
language from this denotation. But first, an obvious remark: any two sets
$C$ and $D$ of cardinality $n$ are always in bijective correspondance. So
we can abstract away from the details of the elements of $C$ and $D$ and
instead choose canonical representations -- in much the same way as computers
choose binary words to represent everything.

\begin{defn} For $n\in\Nat$, denote by $\fin{n}$ the set
$\left\{0,1,\ldots,n-1\right\}$.
We will refer to $\fin{n}$ as the canonical set with $n$ elements.
\end{defn}

Bijections on \fin{n} have a specific name: permutations. As is well-known
permutations can be generated by sequences of transpositions. Thus we
can create a small language for writing permutations as
\begin{verbatim}
p ::= id | swap i j | p;p
\end{verbatim}
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove 
\verb|id|
from the language and drop the $i\neq j$ condition so that 
\verb|swap j j| would represent the identity permutation.

While we are mainly interested in typed data, it is always worthwhile to
first start investigating the untyped setting first, as it gives us insights
into what the \emph{purely operational} view of the theory would be.
This tells us that permutations are an inescapable part of the fabric of
reversible computing.

However, as permutations are untyped and act on the canonicalized version
of $n$-element sets (i.e. those sets where all the structure has been
forgotten), these are a rather pale shadow of the rich tapestry of
information-perserving transformation of structured data.
